
---
title: 'Project 1: Exploratory Data Analysis'
author: "SDS348"
date: ''
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  pdf_document:
    toc: no
---
```{r, echo=FALSE}
knitr::opts_chunk$set(error = TRUE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F,
tidy=TRUE, tidy.opts=list(width.cutoff=60), R.options=list(max.print=100))
```

```{r}
library(ggplot2)
library(ggrepel)
library(tidyr)
library(tidyverse)
library(dplyr)
```

## Ahmed Elsayes (AKE495)
## Data Wrangling and Data Exploration

### Instructions
A knitted R Markdown document (ideally HTML) and the raw R Markdown file (as .Rmd) should both be submitted to Canvas by 11:59pm on 10/11/2020. These two documents will be graded jointly, so they must be consistent (i.e., don’t change the R Markdown file without also updating the knitted document).

The text of the document should provide a narrative structure around your code/output. All results presented must have corresponding code. Any answers/results/plots etc. given without the corresponding R code that generated the result will not be considered. Furthermore, all code contained in your final project document must work correctly (knit early, knit often)! Please do not include any extraneous code or code which produces error messages. (Code that produces warnings is acceptable, as long as you understand what the warnings mean!)

### Find data:

Find two (!) datasets with one variable in common (e.g., dates, times, states, counties, countries, sports players), both with at least 50 observations (i.e., rows) in each. Please think very carefully about whether it makes sense to combine your datasets! If you find one dataset with 50 patients and it has their age, and you find another dataset with 50 *different* patients that has their ages, it makes no sense to join them based on age (you would just be pairing up random people of the same age).

When combined, the resulting/final dataset must have **at least 4 different variables (at least 3 numeric) in addition to the common variable** (i.e., five variables total).

You can have as many variables as you would like! If you found two datasets that you like but they don't have enough variables, find a third dataset with the same common variable and join all three.

```{r}
library(dplyr)
glimpse(X2014) #dataset1
#This dataset was obtained from github and it showed every NBA players' height and weight.
glimpse(stats2) #dataset2
#This dataset was obtained from FiveThirtyEight and it showed every players statistics. There were too many statistics so I tweaked the data and dropped a few unnecessary columns to isolate the points scored statistics.
```
*Introduction: I love sports and I love watching the NBA. The NBA is something I find interesting so I decided to do my project on that. In this project, I have chose two datasets. The first shows NBA player height and weight and the second shows points scored per NBA player. I expect to find an association between weight and point as well as height and points because I think being big and tall gives a huge advantage in basketball because of the leverage you would have. The NBA finals just came to a close with and interesting finish so I decided this was a topic I was most currently interested in. However, these datasets are from the 2014-2015 NBA season. I hope to find interesting correlations and learn something new about the NBA through this project.*


### Guidelines

1. If the datasets are not tidy, you will need to reshape them so that every observation has its own row and every variable its own column. If the datasets are both already tidy, you will make them untidy with `pivot_wider()/spread()` and then tidy them again with `pivot_longer/gather()` to demonstrate your use of the functions. It's fine to wait until you have your descriptives to use these functions (e.g., you might want to pivot_wider() to rearrange the data to make your descriptive statistics easier to look at); it's fine long as you use them at least once!

    - Depending on your datasets, it might be a good idea to do this before joining. For example, if you have a dataset you like with multiple measurements per year, but you want to join by year, you could average over your numeric variables to get means/year, do counts for your categoricals to get a counts/year, etc.
    
    - If your data sets are already tidy, demonstrate the use of `pivot_longer()/gather()` and `pivot_wider()/spread()` on all or part of your data at some point in this document (e.g., after you have generated summary statistics in part 3, make a table of them wide instead of long).

```{r}
#data is already tidy
HeightWeight = subset(X2014, select = -c(3) ) #I dropped the 3rd column which was height in feet and inches because it would cause confusion. Instead, I will just beusing inches for height.
HeightWeight %>% pivot_wider(names_from = Name, values_from = `Height (Inches)`)
#First dataset is now untidy
HeightWeight %>% pivot_longer(everything(), names_to = c(".value", "set"), names_pattern = "(.)(.)") 
glimpse(HeightWeight)
#First dataset is now tidy
stats2 %>% pivot_wider(names_from = Name, values_from = Points) #now untidy
stats2 %>% pivot_longer(everything(), names_to = c(".value", "set"), names_pattern = "(.)(.)")
glimpse(stats2)
#Second dataset is now tidy
```
*The datasets were already tidy so I used the pivot wider function for both datasets and that got them untidy because they were long and that made the datasets wide and so not ever row was assigned to a player and not all cells were filled either. I used the pivot longer function to tidy the datasets back and it worked because each variable has an assigned column, each observation (player) has a row, and there are not empty cells.*

2. Join your 2+ separate data sources into a single dataset based on a common ID variable! If you can't find a good pair datasets to join, you may split one main dataset into two different datasets with a common ID variable in each, and then join them back together based on that common ID, but this is obviously less than ideal.

    - You will document the type of join that you do (left/right/inner/full), including a discussion of how many observations were in each dataset, which observations in each dataset were dropped (if any) and why you chose this particular join. 

```{r}
NBA <- inner_join(HeightWeight, stats2, by="Name")
glimpse(NBA) 
```
*I did an inner join which is basically like a merge and what this did is that it cut all the NBA players out that are not on both datasets. This was necessary because then the merged dataset would have NA values if any other join types were used and those players that were not in both datasets would make it hard to find a correlation and make the dataset messy.*

3. Create summary statistics

    - Use *all six* core `dplyr` functions (`filter, select, arrange, group_by, mutate, summarize`) to manipulate and explore your dataset. For mutate, create a  new variable that is a function of at least one other variable, preferably using a dplyr vector function (see dplyr cheatsheet). It's totally fine to use the `_if`, `_at`, `_all` versions of mutate/summarize instead (indeed, it is encouraged if you have lots of variables)
    
    - Create summary statistics (`mean, sd, var, n, quantile, min, max, n_distinct, cor`, etc) for each of your numeric variables both overall and after grouping by one of your categorical variables (either together or one-at-a-time; if you have two categorical variables, try to include at least one statistic based on a grouping of two categorical variables simultaneously). If you do not have any categorical variables, create one using mutate (e.g., with `case_when` or `ifelse`) to satisfy the `group_by` requirements above. Ideally, you will find a way to show these summary statistics in an easy-to-read table (e.g., by reshaping). (You might explore the kable package for making pretty tables!) If you have lots of numeric variables (e.g., 10+), or your categorical variables have too many categories, just pick a few (either numeric variables or categories of a categorical variable) and summarize based on those. It would be a good idea to show a correlation matrix for your numeric variables (you will need it to make one of your plots).
```{r}
library(dplyr)
NBA %>% group_by(Points) %>% filter(Weight > mean(Weight, na.rm = TRUE)) #Used the group_by function to group by the variable points then I used the filter function to filter for mean weight.
select(NBA, Name, Weight , Points) #Used select function to select three variables. In this case I selected Name, Weight, and Points as my variables.
arrange(NBA, Points , by_group = FALSE) #Used arrange function to arrange the dataset variable points in order from lowest to highest pooints scored.
NBA %>% mutate(rank = min_rank(desc(Weight))) #Used the mutate function to rank weight
NBA %>% group_by(Name) %>% summarize('average height' = mean(`Height (Inches)`, na.rm=T)) #Used the summarize function to ge the average height statistic
NBA %>% group_by(Name) %>% summarize('average weight' = mean(Weight, na.rm=T)) #Used the summarize function to ge the average points scored statistic
simpledata <- NBA[, c(2,3,6)]
sapply(simpledata, is.numeric)
transform(simpledata, Points = as.numeric(Points))
WeightClasses <- NBA %>% mutate(category=cut(Weight, breaks=c(-Inf, 200, 225, Inf), labels=c("low","middle","high"))) #This made three weight classes 
WeightClasses %>% group_by(category) %>% summarize('Average Points' = mean(Points)) 
summarize(group_by(WeightClasses, category), m = mean(Points)) 
HeightClasses <- NBA %>% mutate(category=cut(`Height (Inches)`, breaks=c(-Inf, 75, 80, Inf), labels=c("Short","Medium","Tall")))
HeightClasses %>% group_by(category) %>% summarize('Average Points' = mean(Points))
simpledata <- NBA[, c(2,3,6)]
sapply(simpledata, is.numeric)
simpledata %>% summarize('Average Weight' = mean(Weight)) #summarize mean weight
simpledata %>% summarize('Median weight' = median(Weight)) #summarize median weight
simpledata %>% summarize('Min Weight' = min(Weight)) #summarize minimum weight
simpledata %>% summarize('Max Weight' = max(Weight)) #summarize maximum weight
simpledata %>% summarize('Average Height' = mean(`Height (Inches)`)) #summarize mean height
simpledata %>% summarize('Median Height' = median(`Height (Inches)`)) #summarize median height
simpledata %>% summarize('Min Height' = min(`Height (Inches)`)) #summarize minimum height
simpledata %>% summarize('Max Height' = max(`Height (Inches)`)) #summarize maximum height
```
*I have coded for many summary statistics. These statistics show characteristics for the data and all the variables. From these statistics I can infer that there is no correlation between height and points or weight and points scored. This is contrary to my predictions*
 
4. Make visualizations (three plots)

    -  Make a correlation heatmap of your numeric variables
    -  Create at least two additional plots of your choice with ggplot that highlight some of the more interesting features of your data.
    - Each plot (besides the heatmap) should have at least three variables mapped to separate aesthetics
    - Each should use different geoms (don't do two geom_bars)
    - At least one plot should include `stat="summary"`
    - Each plot should include a supporting paragraph describing the relationships that are being visualized and any trends that are apparent
        - It is fine to include more, but limit yourself to 4. Plots should avoid being redundant! Four bad plots will get a lower grade than two good plots, all else being equal.
    - Make them pretty! Use correct labels, etc.

```{r}
library(reshape2)
library(ggplot2)
simpledata <- NBA[, c(2,3,6)]
cormatrix <- round(cor(simpledata[sapply(simpledata, is.numeric)]),2)
melted <- melt(cormatrix)
get_lower_tri<-function(cormatrix){
    cormatrix[upper.tri(cormatrix)] <- NA
    return(cormatrix)
}

  get_upper_tri <- function(cormatrix){
    cormatrix[lower.tri(cormatrix)]<- NA
    return(cormatrix)
  }
uppertri <- get_upper_tri(cormatrix)
melted2 <- melt(uppertri, na.rm = TRUE)
library(ggplot2)
ggplot(data = melted2, aes(Var2, Var1, fill = value))+
geom_tile(color = "white")+
scale_fill_gradient2(low = "white", high = "red", mid = "pink", 
midpoint = 0, limit = c(-1,1), space = "Lab", 
name="Correlation") +
theme_minimal()+ 
theme(axis.text.x = element_text(angle = 45, vjust = 1, 
size = 12, hjust = 1))+
coord_fixed()
```
*This ggplot shows no visual correlation between height and points for NBA players. This means that there are low point scorers and high point scorers for every height class. This is interesting and it disproves my hypothesis that points scored and height are correlated. This is a scatterplot so it does not show any clear indicators of correlation as opposed to a correlation heatmap which is a stronger indicator.*
```{r}
ggplot(data = NBA, aes(x = `Height (Inches)`, y = Points, color = Name)) + geom_point() + scale_x_continuous(breaks = seq(0,3000,by = 100)) + ggtitle("Points per Player Height")  + labs(y="Points", x = "Height (Inches)") + theme(legend.position = "none") + geom_smooth(aes(group = 1), size = 2, method = "lm", se = FALSE) 
```
*This ggplot shows no visual correlation between height and points for NBA players. This means that there are low point scorers and high point scorers for every height class. This is interesting and it disproves my hypothesis that points scored and height are correlated. This is a scatterplot so it does not show any clear indicators of correlation as opposed to a correlation heatmap which is a stronger indicator.*
```{r}
ggplot(data = WeightClasses, aes(x = category, y = Points, 
    fill = category)) + labs(title = "Average Points Per Weight Class", 
    x = "Weight Class", y = "Points") + geom_errorbar(stat = "summary", 
    width = 0.5) + scale_y_discrete(name = "Points") + 
    geom_point(stat = "summary", size = 10, color = "red")
```
*This plot shows points scored by weight class. I grouped NBA players into three weight classes and created this plot in order to visualize correlations between points and weight. This plot shows that there is no correlation between points and weight. This is opposing my hypothesis because I thought player with more weight would score more points. This is probably because small player such as point gaurds are better at shooting in general so it makes up for their lack of ability to get into the paint and score.*

5. Perform k-means/PAM clustering or PCA on (at least) your numeric variables.

    - Include all steps as we discuss in class, including a visualization.

    - If you don't have at least 3 numeric variables, or you want to cluster based on categorical variables too, convert them to factors in R, generate Gower's dissimilarity matrix on the data, and do PAM clustering on the dissimilarities.
    
    - Show how you chose the final number of clusters/principal components 
    
    - Interpret the final clusters/principal components 

    - For every step, document what your code does (in words) and what you see in the data!     

<P style="page-break-before: always">
\newpage
    
### Rubric

Prerequisite: Finding appropriate data from at least two sources per the instructions above: Failure to do this will result in a 0! You will submit a .Rmd file and a knitted document (pdf).

#### 0. Introduction (4  pts)

- Write a narrative introductory paragraph or two describing the datasets you have chosen, the variables they contain, how they were acquired, and why they are interesting to you. Expand on potential associations you may expect, if any.

#### 1. Tidying: Rearranging Wide/Long (8 pts)

- Tidy the datasets (using the `tidyr` functions `pivot_longer`/`gather` and/or `pivot_wider`/`spread`) 
- If you data sets are already tidy, be sure to use those functions somewhere else in your project
- Document the process (describe in words what was done per the instructions)
    
#### 2. Joining/Merging (8 pts)

- Join your datasets into one using a `dplyr` join function
- If you have multiple observations on the joining variable in either dataset, fix this by collapsing via summarize
- Discuss the process in words, including why you chose the join you did
- Discuss which cases were dropped, if any, and potential problems with this

#### 3. Wrangling (40 pts)

- Use all six core `dplyr` functions in the service of generating summary statistics (18 pts)
    - Use mutate at least once to generate a variable that is a function of at least one other variable

- Compute at least 10 summary statistics for using summarize and summarize with group_by (18 pts)
    - Use at least 5 unique functions inside of summarize (e.g., mean, sd)
    - At least 2 of these should group by a categorical variable. Create one by dichotomizing a numeric if necessary
    - If applicable, at least 1 of these should group by two categorical variables
    - Strongly encouraged to create a correlation matrix with `cor()` on your numeric variables

- Summarize/discuss all results in no more than two paragraphs (4 pts)


#### 4. Visualizing (30 pts)

- Create a correlation heatmap of your numeric variables

- Create two effective, polished plots with ggplot

    - Each plot should map 3+ variables to aesthetics 
    - Each plot should have a title and clean labeling for all mappings
    - Change at least one default theme element and color for at least one mapping per plot
    - For at least one plot, add more tick marks (x, y, or both) than are given by default
    - For at least one plot, use the stat="summary" function
    - Supporting paragraph or two (for each plot) describing the relationships/trends that are apparent
    
#### 5. Dimensionality Reduction (20 pts) 

- Either k-means/PAM clustering or PCA (inclusive "or") should be performed on at least three numeric variables in your dataset

    - All relevant steps discussed in class 
    - A visualization of the clusters or the first few principal components (using ggplot2)
    - Supporting paragraph or two describing results found 


#### 6. Neatness!

- Your project should not knit to more than 30 or so pages (probably closer to 10)! You will lose points if you print out your entire dataset(s), have terrible formatting, etc. If you start your project in a fresh .Rmd file, you are advised to paste the set-up code from this document (lines 14-17) at the top of it: this will automatically truncate if you accidentally print out a huge dataset, etc. Imagine this is a polished report you are giving to your PI or boss to summarize your work researching a topic.


### Where do I find data?

OK, brace yourself!

You can choose ANY datasets you want that meet the above criteria for variables and observations. I'm just sitting here but off the top of my head, if you are into amusement parks, you could look at amusement-park variables, including ticket sales per day etc.; then you could join this by date in weather data. If you are interested in Game of Thrones, you could look at how the frequency of mentions of character names (plus other character variables) and the frequency of baby names in the USA...You could even take your old Biostats data and merge in new data (e.g., based on a Google forms timestamp).

You could engage in some "me-search": You can [request your Spotify data](https://support.spotify.com/ca-en/article/data-rights-and-privacy-settings/) or [download Netflix viewing activity](https://help.netflix.com/en/node/101917), Amazon purchase history, etc. You can use your Google Fit/Fitbit/Apple watch data, etc. These can be combined (e.g., with each other, with other data sources).

You can make it as serious as you want, or not, but keep in mind that you will be incorporating this project into a portfolio webpage for your final in this course, so choose something that really reflects who you are, or something that you feel will advance you in the direction you hope to move career-wise, or something that you think is really neat. On the flip side, regardless of what you pick, you will be performing all the same tasks, so it doesn't end up being that big of a deal.

If you are totally clueless and have no direction at all, log into the server and type 

```{R eval=F}
data(package = .packages(all.available = TRUE))
```

This will print out a list of **ALL datasets in ALL packages** installed on the server (a ton)! Scroll until your eyes bleed! Actually, do not scroll that much... To start with something more manageable, just run the command on your own computer, or just run `data()` to bring up the datasets in your current environment. To read more about a dataset, do `?packagename::datasetname`. 

If it is easier for you, and in case you don't have many packages installed, a list of R datasets from a few common packages (also downloadable in CSV format) is given at the following website:   (including types/numbers of variables in each)

- A good package to download for fun/relevant data is `fivethiryeight`. Just run `install.packages("fivethirtyeight"), load the packages with `library(fivethirtyeight)`, run `data()`, and then scroll down to view the datasets. Here is an online list of all 127 datasets (with links to the 538 articles). Lots of sports, politics, current events, etc: https://cran.r-project.org/web/packages/fivethirtyeight/vignettes/fivethirtyeight.html

- If you have already started to specialize (e.g., ecology, epidemiology) you might look at discipline-specific R packages (vegan, epi, respectively). We will be using some tools from these packages later in the course, but they come with lots of data too, which you can explore according to the directions above

- However, you *emphatically DO NOT* have to use datasets available via R packages! In fact, I would much prefer it if you found the data from completely separate sources and brought them together (a much more realistic experience in the real world)! You can even reuse data from your SDS328M project, provided it shares a variable in common with other data which allows you to  merge the two together (e.g., if you still had the timestamp, you could look up the weather that day: https://www.wunderground.com/history/). If you work in a research lab or have access to old data, you could potentially merge it with new data from your lab!

- Here is a curated list of interesting datasets (read-only spreadsheet format): https://docs.google.com/spreadsheets/d/1wZhPLMCHKJvwOkP4juclhjFgqIY8fQFMemwKL2c64vk/edit

- Here is another great compilation of datasets: https://github.com/rfordatascience/tidytuesday

- Here is the UCI Machine Learning Repository: https://archive.ics.uci.edu/ml/index.php

    - See also https://en.wikipedia.org/wiki/List_of_datasets_for_machine-learning_research#Biological_data

- Here is another good general place to look: https://www.kaggle.com/datasets

- To help narrow your search down or to see interesting variable ideas, check out https://www.tylervigen.com/spurious-correlations. This is the spurious correlations website, and it is fun, but if you look at the bottom of each plot you will see sources for the data. This is a good place to find very general data (or at least get a sense of where you can scrape data together from)!

- If you are interested in medical data, check out www.countyhealthrankings.org

- If you are interested in scraping UT data, the university makes *loads* of data public (e.g., beyond just professor CVs and syllabi). Check out all the data that is available in the statistical handbooks: https://reports.utexas.edu/statistical-handbook

##### Broader data sources:

[Data.gov](www.data.gov) 186,000+ datasets!

[Social Explorer](Social Explorer) is a nice interface to Census and American Community Survey data (more user-friendly than the government sites). May need to sign up for a free trial.

[U.S. Bureau of Labor Statistics](www.bls.gov)

[U.S. Census Bureau](www.census.gov)

[Gapminder](www.gapminder.org/data), data about the world.

...





